### 2017-09-19 marking_response/response.tex

\noindent
\textcolor{blue}{
	\textit{A reasonably comprehensive literature review was provided, however of concern, over 8.5 pages of the introduction is effectively a regurgitation of the `Hallmarks of cancer' review by Weinberg and Hanahan.
	This could be substantially cropped.
	}
}\\

\noindent
\\

\noindent
\textcolor{blue}{
	\textit{Given this thesis is focused on a known environmental influence on cancer (obesity), it was a surprise the section on environmental factors associated with cancer incidence (1.2.3) is underdeveloped and superficial.
	}
}\\

\noindent
I understand your concern about the relatively brief description of the known environmental factors that affect cancer incidence.
As you may know, the focus of this project was the relationship between cancer and obesity, and how obesity affects cancer (which is covered more in depth in Section 1.3.2).

@@

Section 1.2 was written as a section introducing cancer in general, and was not specifically focused on any single cancer type, nor environmental factor.
So it follows that Section 1.2.3 describes, in very general terms, how \textit{some} (not \textit{all}) environmental factors can influence the cancer incidence, and I think the amount of information placed in this subsection is sufficient for this purpose.
Additionally, the environmental factors chosen to be included in this section was based on either (1) (very) strong evidence of their role in cancer onset/progression in the literature (in terms of both epidemiological and biological evidence/association), or (2) some sort of relationship with obesity.

Perhaps, on request, I could emphasise these points within Section 1.2.3 and guide the readers to Section 1.3.2 for specific details on the effect of obesity on cancer.
\\

\noindent
\textcolor{blue}{
	\textit{In Section 1.2.4, Provide a citation for a recent list of molecular targets in cancer therapy, and their companion approved targeted agents (or present it in a table in appendix).
	This first paragraph is superficial and na\"ive.
	What is the rationale for only mentioning the RAS/RAF/MEK/ERK and PI3K/AKT pathways?
	Is it because these are the most relevant to those impacted by obesity?
	It would follow that those pathways be more in alignment with inflammatory signalling pathways rather than those associated with cell proliferation (as mentioned later in section 1.3.2).
	The rationale for cherry picking select cell signalling information for this section of the introduction is unclear.
	}
}\\

\noindent
As mentioned above, Section 1.2 was intended as a general overview of cancer, and not specific to any particular environmental factor or cancer type.
RAS-ERK pathway was chosen as an example of the target for cancer treatment, as this signalling pathway was arguably the most well-studied and understood signalling pathway in cancer.
PI3K/Akt pathway was included very briefly as an example of how complex the biology of a tumour could be, and the associated difficulties in treating cancer.
\\

@@

\noindent
\textcolor{blue}{
\textit{``\ldots none of the metagenes were not generalisable in other cancer datasets.'' does not make grammatical sense (p83)
}
}\\

\noindent
Changed ``\ldots none of the metagenes were not generalisable in other cancer datasets'' to ``\ldots none of the metagenes were generalisable in other cancer datasets''
\\

\noindent
\textcolor{blue}{
\textit{The font used in the plots for figures 19-22 are exceedingly small (p100, 106)
}
}\\

\noindent
I agree that the fonts are indeed small, but due to the nature of the plot arrangement for these figures the fonts cannot be made any larger (at least not trivially).
Also, the font size was chosen so that the statistical values and letters did not obscure the points and boxes of the plots, and I would like to avoid this as much as possible, so I have left the font size unchanged.
\\

@@

\newpage

\section*{Response to the comments made by Examiner ``1137''}
\label{sec:response_to_comments_from_examiner_1137}

\noindent
\textcolor{blue}{
	\textit{Individually published papers could have been critically appraised in the Introduction when it was introduced.
	}
}\\

\noindent
Slightly more details on Creighton \textit{et al.} and Fuentes-Mattei \textit{et al.} papers have been added in Section 1.4.
\\

@@

\noindent
\textcolor{blue}{
	\textit{A general in-depth exploration of the `meaning' of the different data types used in the thesis and how they are affected by normalisation and statistical test choice would have been appropriate in the Introduction.
	}
}\\

\noindent
\\

\noindent
\textcolor{blue}{
	\textit{How does the overlap between the genes that constitute each metagene influence the statistical analysis?
	}
}\\

\noindent
I have not looked into the overlap between the genes/probes between different genetic signatures in great details.
The overlap of the genes between the different signatures will make it more/less similar to one another (depending on the overlap).
Since there were no direct comparison between any two genetic signatures, I doubt that there was any influence on the statistical analysis carried out in this project.
Note that multiple genetic signatures were used in the linear models as covariates for controlling their influence on the models that were constructed, and therefore it would have had no negative influence on the analyses.
\\

@@

\noindent
From Table 2, the NZBC and the Gatza \textit{et al.} data sets were the only data set where the samples were excluded.
For NZBC data set, only the samples that had BMI or obesity status  information was included in the analysis (not stated in the thesis --  will be included after revision).
For Gatza \textit{et al.} data set, only the samples that were genotyped with Affymetrix HGU133A platform was used in the analysis, as all of the other three microarray data sets used this platform.

These information were stated in Section 2.2.1 under the corresponding data sets, but perhaps it might have been better if it was stated separately in a section of its own.
\\

\noindent
\textcolor{blue}{
	\textit{Why was the BRCA RNAseq data set in TCGA not used?
	Surely this could have been informative?
	}
}\\

\noindent
I agree with the Examiner and the BRCA data set would have aided greatly in the exploration of obesity-associated signatures from both Creighton \textit{et al.} and Fuentes-Mattei \textit{et al.} data sets.
However, at the time of data retrieval (March 31st 2015), there were no height, weight or BMI information for the BRCA data set, and could not use the data set for the analyses.
\\

@@

\noindent
\textcolor{blue}{
	\textit{Cancer patients can have vast differences in BMI at different points in their treatment.
	Was the clinical timing of weight measures standardised in some way to reduce the impact of this problem, e.g. using weight at initial diagnosis?
	}
}\\

\noindent
This is a good point, and only Creighton \textit{et al.} have specifically stated in their paper that the patient BMI was measured at the time of diagnosis.
Fuentes-Mattei \textit{et al.} paper did not describe when the patient BMI was measured.
Although I am not aware of how and when clinicians record patient data, I would suspect they would have measured BMI at the time of diagnosis to keep the data consistent with other variables (e.g. tumour grade).
\\

\noindent
\textcolor{blue}{
	\textit{In the Introduction, add a few sentences of additional explanation about potential bias and confounding introduced into previously published studies of obesity-cancer gene expression relationships, including potential confounding by lifestyle factors associated with obesity.
	}
}\\

\noindent
This has been included in the added paragraph in Section 1.2.3.
\\

@@

\noindent
\textcolor{blue}{
	\textit{In the Introduction, refer to the published in vitro studies of the effect on cancer cells of factors produced by adipocytes in culture.
	}
}\\

\noindent
\\

\noindent
\textcolor{blue}{
	\textit{In the Introduction, add a few additional sentences to more fully introduce potential mechanisms at the level of cancer cells for obesity to alter biology through changes in hormone receptor expression and transcriptional effects of endocrine factors like leptin and the IGFs.
	}
}\\

\noindent
Added a couple of sentences at the end of second paragraph in Section 1.3.2.
\\

@@

\noindent
\textcolor{blue}{
	\textit{Add additional description about what basic quality control was possible for each data set (perhaps in section 2.3), and the results of the QC.
	The discussion in section 5.2.2 is hard to interpret without this.
	}
}\\

\noindent
\\

\noindent
\textcolor{blue}{
	\textit{It would be interesting to explore whether the differences in risk ratio between cancer types was due to technical issues, or represented real biology.
	Discuss whether the biology underpinning differences between cancer types in terms of obesity risk ratios in Table 1 could be explored statistically using genomic data.
	}
}\\

\noindent
Though there would have been some technical issues in the individual studies included in the meta-analysis study by Renehan \textit{et al.}, I suspect that there is an effect of obesity on certain types of cancer.
I am not sure whether there are any genomic data that could clarify the Examiner's question, as it is possible that obesity affects many different biological functions/pathways of a cell.
However, I wonder if it would be possible to use SVD, or some other weighting scores, that take these risk ratios into account and essentially ``normalise'' the genetic signals (e.g. gene expression and/or methylation) across different cancer types, and thereby focus on the underlying biology of each cancer type.
\\

### 2017-09-19 thesis/intro/intro.tex
Final edit

In addition to the mechanisms described above, there is a growing body of evidence where some environmental factors may affect the transcriptional regulation of a set of genes in certain cancer type.
A relatively recent review by \citet{Perez-Solis2016} describes potential mechanism in which obesity, alcoholism and smoking may regulate gene expression in breast cancer cells.
By affecting the so-called ``Master'' transcription factors, environmental factors are able to affect the expression of the genes that are downstream to the Master transcription factor, and as a result, lead to the development of cancerous cells \citep{Perez-Solis2016}.
It is not surprising that different environmental factors affect different Master transcriptional factor, and therefore have different biological mechanism of gene regulation.

@@

Adipose tissue has long known to have been a group of cells to store excess fats, but it has recently become apparent that adipose tissue is also a metabolically active endocrine organ \citep{Roberts2010}.
Increased adiposity is known to influence the bioavailability of endogenous sex hormones, including estrogens, androgens and progesterone \citep{Calle2004}.
\Glspl{adipocyte} synthesise enzymes, such as aromatase, that convert precursor molecules into different forms of androgens and estrogens \citep{Calle2004}.
Many studies have confirmed the mitogenic effect of sex hormones on certain cell types, mainly breast and endometrial cell types \citep{Roberts2010}.

On top of the effects that \glspl{adipocyte} have on circulating concentration of sex hormones, \glspl{adipocyte} also produce adipose tissue-specific hormones called adipokines \citep{Roberts2010}.
Of the many adipokines identified so far, leptin and adipo\-nectin have been studied the most in terms of cancer development \citep{Renehan2006,Roberts2010}.
Leptin is known to activate not only the satiety signalling pathway, but also the \gls{jak}/\gls{stat} and \gls{pi3k}/Akt signalling pathways \citep{Garofalo2006,Renehan2006}.
Activation of these pathways leads to the activation of the target genes that are involved in cell proliferation and cell survival, as well as pro-angiogenic factors \citep{Garofalo2006}.

@@

\citet{Creighton2012} used 103 breast tumour samples and analysed these samples using microarrays.
Of the 103 samples, 35\% were from normal weight, 28\% were from overweight and 37\% were from obese patients.
It was also noted that African-American patients were more likely to be obese than Caucasian patients.
They compared the gene expression between the tumours from obese patients with the tumours from non-obese patients and identified 799 gene probe sets (662 unique genes) with p \textless{} 0.01 and with a \gls{fc} \textgreater{} 1.2.
The majority of these genes (725 gene probes) were downregulated and 74 gene probes were upregulated in the breast tumour samples from the obese patients compared with the breast tumour samples from the non-obese patients.

@@

\citet{Fuentes-Mattei2014} studied 137 breast tumour samples from \gls{er} positive patients and analysed their transcriptome using microarrays.
Of the 137 patients, 43 patients were obese and 94 patients were not obese.
Comparison of the gene expressions of the tumour samples from the obese patients with the tumour samples from the non-obese patients revealed 112 genes that were statistically significantly dysregulated (62 genes upregulated and 50 genes downregulated).
Furthermore, functional transcriptomic analysis suggested that the insulin signalling and inflammation affected human \gls{er}$^+$ breast cancer.

On top of this, \citet{Fuentes-Mattei2014} validated their results using a mouse model, where they compared the expression level in the breast tumours from the obese mice with those from the lean mice.
Results from this animal model provided strong evidence that obesity significantly accelerated tumour growth through Akt/\gls{mtor} signalling pathway.

>>>>

In addition to the mechanisms described above, there is a growing body of evidence where some environmental factors may affect the transcriptional regulation of a set of genes in certain cancer type.
A relatively recent review by \citet{Perez-Solis2016} describes potential mechanism in which obesity, alcoholism and smoking (all a lifestyle factor related to obesity) may regulate gene expression in breast cancer cells.
By affecting the so-called ``Master'' transcription factors, environmental factors are able to affect the expression of the genes that are downstream to the Master transcription factor, and as a result, lead to the development of cancerous cells \citep{Perez-Solis2016}.
It is not surprising that different environmental factors affect different Master transcriptional factor, and therefore have different biological mechanism of gene regulation.

@@

Adipose tissue has long known to have been a group of cells to store excess fats, but it has recently become apparent that adipose tissue is also a metabolically active endocrine organ \citep{Roberts2010}.
Increased adiposity is known to influence the bioavailability of endogenous sex hormones, including estrogens, androgens and progesterone \citep{Calle2004}.
\Glspl{adipocyte} synthesise enzymes, such as aromatase, that convert precursor molecules into different forms of androgens and estrogens \citep{Calle2004}.
Many studies have confirmed the mitogenic effect of sex hormones on certain cell types, mainly breast and endometrial cell types \citep{Roberts2010}.
In addition to this, there has also been evidence of serum from obese patient influencing the gene expression of sex hormone receptor (estrogen receptor $\beta$) in a breast cancer cell line, suggesting there may be a gene regulatory effect caused by the obese phenotype \citep{Bowers2015}.
Through the effects of dysregulated levels of hormones and hormone receptors, obesity may have significant impact on the tumour biology through changes in gene expression level.

On top of the effects that \glspl{adipocyte} have on circulating concentration of sex hormones as well as gene expression of sex hormone receptors, \glspl{adipocyte} also produce adipose tissue-specific hormones called adipokines \citep{Roberts2010}.
Of the many adipokines identified so far, leptin and adipo\-nectin have been studied the most in terms of cancer development \citep{Renehan2006,Roberts2010}.
Leptin is known to activate not only the satiety signalling pathway, but also the \gls{jak}/\gls{stat} and \gls{pi3k}/Akt signalling pathways \citep{Garofalo2006,Renehan2006}.
Activation of these pathways leads to the activation of the target genes that are involved in cell proliferation and cell survival, as well as pro-angiogenic factors \citep{Garofalo2006}.

@@

\citet{Creighton2012} used 103 breast tumour samples and analysed these samples using microarrays.
The tumour samples were trimmed upon collection, to ensure all of the samples had $\geq$ 70\% tumour cells before \acrshort{rna} was extracted from the samples.
Of the 103 samples, 35\% were from normal weight, 28\% were from overweight and 37\% were from obese patients.
It was also noted that African-American patients were more likely to be obese than Caucasian patients.
They compared the gene expression between the tumours from obese patients with the tumours from non-obese patients and identified 799 gene probe sets (662 unique genes) with p \textless{} 0.01 and with a \gls{fc} \textgreater{} 1.2.
The majority of these genes (725 gene probes) were downregulated and 74 gene probes were upregulated in the breast tumour samples from the obese patients compared with the breast tumour samples from the non-obese patients.

Major concern with this study was the quality of the \acrshort{rna} that was collected from the samples.
Ensuring that all of the samples had more than 70\% of tumour cells meant that the \acrshort{rna} was likely from the tumour cells.
However, this did not control for the quality of the \acrshort{rna} that has been extracted, as some samples may have had significantly less tumour volume compared to another, but still have $\geq$ 70\% tumour cells in both samples.
Another thing to note from this paper was the predictive ability of their obesity signatures in other data sets.
\citet{Creighton2012} used six independent breast cancer  data sets to test whether their signature was able to replicate their findings.
Although they were able to replicate their results in one of the data sets, they were not able to find any significant association in the other data sets.

@@

\citet{Fuentes-Mattei2014} studied 137 breast tumour samples from \gls{er} positive patients and analysed their transcriptome using microarrays.
Of the 137 patients, 43 patients were obese and 94 patients were not obese.
Comparison of the gene expressions of the tumour samples from the obese patients with the tumour samples from the non-obese patients revealed 112 genes that were statistically significantly dysregulated (62 genes upregulated and 50 genes downregulated).
Furthermore, functional transcriptomic analysis suggested that the insulin signalling and inflammation affected human \gls{er}$^+$ breast cancer.
One minor thing to point out in this paper was that the normalisation procedure carried out before the differential expression analysis was not clear.

On top of the bioinformatic analysis that was carried out, \citet{Fuentes-Mattei2014} validated their results using a mouse model, where they compared the expression level in the breast tumours from the obese mice with those from the lean mice.
Results from this animal model provided strong evidence that obesity significantly accelerated tumour growth through Akt/\gls{mtor} signalling pathway.

### 2017-09-19 thesis/methods/methods.tex
Final edit

\begin{table}[htpb]
	\centering
	\caption{Number of samples in each of the breast cancer microarray data}
	\label{tab:num_sample_microarray}
	\begin{tabular}{lc}
		Data set & Number of samples\\
		\hline
		\hline
		\rule{0pt}{2.25ex}Creighton \textit{et al.} & 103\\
		Fuentes-Mattei \textit{et al.}              & 278\\
		\gls{nzbc}                                  & 99 \\
		Gatza \textit{et al.}                       & 1060\\
		\hline
		\hline
	\end{tabular}
\end{table}

>>>>

\begin{table}[htpb]
	\centering
	\caption{Number of samples in each of the breast cancer microarray data}
	\label{tab:num_sample_microarray}
	\begin{threeparttable}
		\begin{tabular}{lcc}
			Data set & Number of samples before exclusion\tnote{1} & Number of samples\\
			\hline
			\hline
			\rule{0pt}{2.25ex}Creighton \textit{et al.} & 103 & 103\\
			Fuentes-Mattei \textit{et al.}              & 278 & 278\\
			\gls{nzbc}                                  & 107 & 99 \\
			Gatza \textit{et al.}                       & 1965 & 1060\\
			\hline
			\hline
		\end{tabular}
		\begin{tablenotes}
			\begin{footnotesize}
			\item [1] See \cref{sub:breast_cancer_data} for exclusion criteria.
			\end{footnotesize}
		\end{tablenotes}
	\end{threeparttable}
\end{table}

### 2017-09-19 thesis/misc/appendix2.tex
Final edit

>>>>

	\section{Overlap of the genes between the pathway-associated signatures}
	\label{sec:overlap_of_the_genes_between_the_pathway_associated_signatures}

	\begin{figure}[htpb]
		\centering
		\includegraphics[width=0.8\linewidth,page=1]{results2/overlap}
		\caption[Heatmap showing the overlap of genes between the different pathway-associated signatures]{Heatmap showing the overlap of genes between the different pathway-associated signatures. For each pairs of pathway-associated signatures, the overlap of genes were counted and divided by the total number of unique genes from the two signatures to get a ratio. This ratio was then scaled and plotted as shown}
		\label{fig:overlap_pathway_genetic_signature_extra}
	\end{figure}

### 2017-09-19 thesis/results/results2.tex
Final edit

In \cref{tab:svd_vs_tm_path,tab:svd_vs_tm_obs}, the obesity and pathway-associated genetic signatures had a correlation of 1 in CR and GT data sets respectively, as the obesity and pathway TMs were generated in the CR and GT data sets, respectively.
The correlations of the obesity and pathway metagenes were variable across different data sets, which was as expected since all the data sets were different from one another.
However, what was unexpected was the fact that there were some pathway genetic signatures that were highly correlated in all of the data sets, and there were others that varied considerably across different data sets.
For example, the \gls{bcat} pathway metagene was highly correlated in all of the data sets (\textgreater{} 0.98), whereas the \gls{stat3} pathway metagene was variable across different data sets, ranging from 0.1902 to 0.9159 (\cref{tab:svd_vs_tm_path}; \cref{sec:comparison_of_the_correlation_of_svd_and_tm_generated_pathway_metagenes}).
The fact that some pathway-associated metagenes were consistent across different data sets suggested that some of these genetic signatures were reliable and did not depend on the data set the transformation matrices were derived from.
On the other hand, the genetic signatures that were not consistent across the data sets were likely to be dependent on the data set in which they were derived from, and therefore the transformation matrices for these signatures must be derived from the GT data set.
This also showed that the tumour biology may be different across the data sets, likely due to the difference in the patient cohorts that were selected for each of the studies.

>>>>

In \cref{tab:svd_vs_tm_path,tab:svd_vs_tm_obs}, the obesity and pathway-associated genetic signatures had a correlation of 1 in CR and GT data sets respectively, as the obesity and pathway TMs were generated in the CR and GT data sets, respectively.
The correlations of the obesity and pathway metagenes were variable across different data sets, which was as expected since all the data sets were different from one another.
However, what was unexpected was the fact that there were some pathway genetic signatures that were highly correlated in all of the data sets, and there were others that varied considerably across different data sets.
For example, the \gls{bcat} pathway metagene was highly correlated in all of the data sets (\textgreater{} 0.98), whereas the \gls{stat3} pathway metagene was variable across different data sets, ranging from 0.1902 to 0.9159 (\cref{tab:svd_vs_tm_path}; \cref{sec:comparison_of_the_correlation_of_svd_and_tm_generated_pathway_metagenes}).
The fact that some pathway-associated metagenes were consistent across different data sets suggested that some of these genetic signatures were reliable and did not depend on the data set the transformation matrices were derived from.
Adding to this, there were very little overlap in the genes between the different pathway-associated genetic signatures, suggesting that the consistency across the data sets were not due to an artefact that was affecting the dominant probe set in the signatures (\cref{sec:overlap_of_the_genes_between_the_pathway_associated_signatures}).
On the other hand, the genetic signatures that were not consistent across the data sets were likely to be dependent on the data set in which they were derived from, and therefore the transformation matrices for these signatures must be derived from the GT data set.
This also showed that the tumour biology may be different across the data sets, likely due to the difference in the patient cohorts that were selected for each of the studies.
